### 2021-06-01 cued/inference.tex
Proof-read night before

We adopt the Markov chain Monte Carlo procedure of
\cite{Peixoto-MCMC},
which relies on writing the posterior in the following form:
%
\begin{equation}
	p(b | A, X) \propto p(A | b, X) \cdot p(b | X) = \pi_b(b).
\end{equation}
%
Now $\pi_b(\cdot)$ is the un-normalised target density,
which can be expressed as:
%  we wish to sample from for the $b$-chain. In other words, we 
% wish to construct a Markov chain that has $\pi_b(\cdot)$ 
% as its invariant distribution. We can break $\pi_b$ down as follows:
%
\begin{equation}
	\pi_b(b) = p(b|X) \sum_{\psi} \nolimits p(A , \psi | b, X) \\
	= p(b|X) p(A, \psi^* | b, X) \\
	= p(A | b, \psi^*) \cdot p(\psi^* | b) \cdot p(b | X).
\end{equation}
%
Since we are using the microcanonical SBM formulation, there is only one 
value of $\psi$ that is compatible with the given $(A, b)$ pair;
recall the constraints in~(\ref{eqn:sbm-constraints}). 
We denote this value $\psi^* = \{\psi_k^*, \psi_e^*\}$. Therefore, 
the summation over all $\psi$ reduces to just the single $\psi^*$ term.
% ; this is the power of the microcanonical formulation. 
We also define the microcanonical entropy of the configuration as,
%
\begin{equation}
	S(b) = - \log \pi_b(b) = - \Big( \log p(A | b, \psi^*) + \log p(\psi^*, b | X) \Big).
	\label{eqn:dl-form}
\end{equation}
%
This entropy can be thought of as the optimal
``description length'' of the graph. 
The exact form of the proposal $q_b$ is explored thoroughly in
\cite{Peixoto-MCMC} and not repeated here. There is a widely used 
library for Python made available under LGPL 
called \verb*|graph-tool| \cite{peixoto_graph-tool_2014}, 
which implements this algorithm. The only modification we make is in 
the block membership prior $p(b)$ that we replace with $p(b|X)=B^{-N}$, 
which cancels out in the MH accept-reject step as it is independent of $b$.

@@

See Appendix \ref{appdx:form-U} for the derivation. The function $U(\theta)$ is a typical objective function for neural network training. The first term is introduced by the likelihood; we collect it into $N \cdot \Lcal(\theta)$, which is the cross-entropy between the graph-predicted and feature-predicted block memberships summed over all vertices. 
The second, introduced by the prior, brings a form of regularisation, guarding against over-fitting. Unlike in many applications where the goal 
is to find the minimiser of $U(\theta)$, our goal is to draw samples from the posterior $\pi_\theta(\cdot) \propto \exp(-U(\cdot))$. We can use $\nabla U$ as a useful heuristic to bias our proposal towards regions of higher target density. We therefore adopt the Metropolis-adjusted Langevin algorithm (MALA) \cite{mala-tweedie}. Given the current sample $\theta$, we generate 
a new sample $\theta'$ as,
%
\begin{equation*}
	\theta' = \theta - h \nabla U(\theta) + \sqrt{2h} \cdot \xi,
\end{equation*}
%
where $\xi \sim \Gaussian(0, I)$ and $h$ is a step-size parameter 
which may vary with the sample index (see Appendix \ref{appdx:step-size} for a discussion of step-size).
This leads to the proposal distribution,
\begin{equation*}
	q_\theta(\theta, \theta') \sim \Gaussian \left( \theta' ; \theta - h \nabla U(\theta), 2h I \right).
\end{equation*}
%
Without the injected noise term, MALA is equivalent to gradient descent. We require the noise term $\xi$ to fully explore the parameter space. 
The term $\nabla U$ has an easy to compute analytic form (derived in Appendix \ref{appdx:gradu}). By noting that $\theta = \{w_k\}_{k=1}^{B}$, we write the derivative with respect to each $w_k$ as,
%
\begin{equation}
	\frac{\partial U}{\partial w_k} = - \left( \sum_{i \in [N]} \Big\{ x_i (y_{ik} - a_{ik}) \Big\} - \frac{w_k}{\sigma_\theta^2} \right).
	\label{eqn:U-derivative}
\end{equation}
%
After a proposed move is generated from $q_\theta$,
it is either accepted or rejected in the standard
Metropolis-Hastings accept/reject step.

@@

Once the samples $\left\{ \theta^{(t)} \right\} \sim p(\theta | A, X)$
have been obtained, we can compute the empirical mean and standard deviation of each component of $\theta$. Switching back to matrix notation we define $\theta = W$, such that $W_{ij}$ is the weight component for block $i$ and feature $j$. We can then define:
%
\begin{equation}
	\hat{\mu}_{ij} \coloneqq \frac{1}{|\Tcal_\theta|} \sum_{t \in \Tcal_\theta} W_{ij}^{(t)} \qquad \textrm{and} \qquad
	\hat{\sigma}_{ij}^2 \coloneqq \frac{1}{|\Tcal_\theta|} \sum_{t \in \Tcal_\theta} \left( W_{ij}^{(t)} - \hat{\mu}_{ij} \right)^2.
\end{equation}
%
A simple heuristic to discard the least important features requires specifying a cutoff $c > 0$ and a multiplier $k > 0$. We define the function $\Fcal_i(j)$ 
as in~(\ref{eqn:fij}) and only keep features with indices $d \in \Dcal'$, where $\Dcal'$ is given in~(\ref{eqn:kept-feature-set}).
%
\begin{align}
	\Fcal_i(j) &\coloneqq (\hat{\mu}_{ij} - k \hat{\sigma}_{ij}, \hat{\mu}_{ij} + k \hat{\sigma}_{ij}) \cap (-c, +c),
	\label{eqn:fij} \\
	\Dcal' &\coloneqq \left\{ j \in [D] : \exists i \in [B] \textrm{ s.t. }  \Fcal_i(j) = \emptyset \right\}
	\label{eqn:kept-feature-set}
\end{align}
%
Intuitively, this means discarding any feature $j$ for which 
$(\hat{\mu}_{ij} - k\hat{\sigma}_{ij}, \hat{\mu}_{ij} + k \sigma)$ overlaps with
$(-c, c)$ for all block indices $i$. If we were to use the Laplace approximation for the posterior $p(W_{ij} | A, X) \approx \Gaussian(W_{ij}; \hat{\mu}_{ij}, \hat{\sigma}_{ij}^2)$, then this is analogous to a hypothesis test on the value of $W_{ij}$ as in~(\ref{eqn:hyp-test-discard});
then $\Dcal'$ comprises all features $j$ for which $H_1$ is accepted at least once for some $i \in [B]$.
%
\begin{equation}
	H_0: |W_{ij}| \leq c \qquad
	H_1: |W_{ij}| > c
	\label{eqn:hyp-test-discard}
\end{equation}
%
The multiplier $k$ in~(\ref{eqn:fij}) determines the degree of significance of the result. However, as the Laplace approximation is not exact we will only treat this dimensionality reduction method as a useful heuristic and not an exact method. Appendix~\ref{appdx:hyp-test} explores this subject at greater depth.

>>>>

We adopt the Markov chain Monte Carlo procedure of
\cite{Peixoto-MCMC},
which relies on writing the posterior in the following form:
%
\begin{equation}
	p(b | A, X) \propto p(A | b, X) \cdot p(b | X) = \pi_b(b).
\end{equation}
%
Now $\pi_b(\cdot)$ is the un-normalised target density,
which can be expressed as:
%  we wish to sample from for the $b$-chain. In other words, we 
% wish to construct a Markov chain that has $\pi_b(\cdot)$ 
% as its invariant distribution. We can break $\pi_b$ down as follows:
%
\begin{equation}
	\pi_b(b) = p(b|X) \sum_{\psi} \nolimits p(A , \psi | b, X) \\
	= p(b|X) p(A, \psi^* | b, X) \\
	= p(A | b, \psi^*) \cdot p(\psi^* | b) \cdot p(b | X).
\end{equation}
%
Since we are using the microcanonical SBM formulation, there is only one 
value of $\psi$ that is compatible with the given $(A, b)$ pair;
recall the constraints in~(\ref{eqn:sbm-constraints}). 
We denote this value $\psi^* = \{\psi_k^*, \psi_e^*\}$. Therefore, 
the summation over all $\psi$ reduces to just the single $\psi^*$ term.
% ; this is the power of the microcanonical formulation. 
We also define the microcanonical entropy of the configuration as,
%
\begin{equation}
	S(b) = - \log \pi_b(b) = - \Big( \log p(A | b, \psi^*) + \log p(\psi^*, b | X) \Big).
	\label{eqn:dl-form}
\end{equation}
%
This entropy can be thought of as the optimal
``description length'' of the graph; if we 
constructed an optimal code for the SBM ensemble,
this would be the minimum amount of information 
required to store the graph and its parameters. 
The exact form of the proposal $q_b$ is explored thoroughly in
\cite{Peixoto-MCMC} and not repeated here. There is a widely used 
library for Python made available under LGPL 
called \verb*|graph-tool| \cite{peixoto_graph-tool_2014}, 
which implements this algorithm. The only modification we make is in 
the block membership prior $p(b)$ that we replace with $p(b|X)=B^{-N}$, 
which cancels out in the MH accept-reject step as it is independent of $b$.

@@

See Appendix \ref{appdx:form-U} for the derivation. The function $U(\theta)$ is a typical objective function for neural network training. The first term is introduced by the likelihood; we collect it into $N \cdot \Lcal(\theta)$, which is the cross-entropy between the graph-predicted and feature-predicted block memberships summed over all vertices. 
The second, introduced by the prior, brings a form of regularisation, guarding against over-fitting. Unlike in many applications where the goal 
is to find the minimiser of $U(\theta)$, our goal is to draw samples from the posterior $\pi_\theta(\cdot) \propto \exp(-U(\cdot))$. We can use $\nabla U$ as a useful heuristic to bias our proposal towards regions of higher target density. We therefore adopt the Metropolis-adjusted Langevin algorithm (MALA) \cite{mala-tweedie}. Given the current sample $\theta$, we generate 
a new sample $\theta'$ as,
%
\begin{equation*}
	\theta' = \theta - h \nabla U(\theta) + \sqrt{2h} \cdot \xi,
\end{equation*}
%
where $\xi \sim \Gaussian(0, I)$ and $h$ is a step-size parameter 
which may vary with the sample index (see Appendix \ref{appdx:step-size} for a discussion of step-size variation).
This leads to the proposal distribution,
\begin{equation*}
	q_\theta(\theta, \theta') \sim \Gaussian \left( \theta' ; \theta - h \nabla U(\theta), 2h I \right).
\end{equation*}
%
Without the injected noise term, MALA is equivalent to gradient descent. We require the noise term $\xi$ to fully explore the parameter space. 
The term $\nabla U$ has an easy to compute analytic form (derived in Appendix \ref{appdx:gradu}). By noting that $\theta = \{w_k\}_{k=1}^{B}$, we write the derivative with respect to each $w_k$ as,
%
\begin{equation}
	\frac{\partial U}{\partial w_k} = - \left( \sum_{i \in [N]} \Big\{ x_i (y_{ik} - a_{ik}) \Big\} - \frac{w_k}{\sigma_\theta^2} \right).
	\label{eqn:U-derivative}
\end{equation}
%
After a proposed move is generated from $q_\theta$,
it is either accepted or rejected in the standard
Metropolis-Hastings accept/reject step.

@@

Once the samples $\left\{ \theta^{(t)} \right\} \sim p(\theta | A, X)$
have been obtained, we can compute the empirical mean and standard deviation of each component of $\theta$. Switching back to matrix notation we define $\theta = W$, such that $W_{ij}$ is the weight component for block $i$ and feature $j$. We can then define:
%
\begin{equation}
	\hat{\mu}_{ij} \coloneqq \frac{1}{|\Tcal_\theta|} \sum_{t \in \Tcal_\theta} W_{ij}^{(t)} \qquad \textrm{and} \qquad
	\hat{\sigma}_{ij}^2 \coloneqq \frac{1}{|\Tcal_\theta|} \sum_{t \in \Tcal_\theta} \left( W_{ij}^{(t)} - \hat{\mu}_{ij} \right)^2.
\end{equation}
%
A simple heuristic to discard the least important features requires specifying a cutoff $c > 0$ and a multiplier $k > 0$. We define the function $\Fcal_i(j)$ 
as in~(\ref{eqn:fij}) and only keep features with indices $d \in \Dcal'$, where $\Dcal'$ is given in~(\ref{eqn:kept-feature-set}).
%
\begin{align}
	\Fcal_i(j) &\coloneqq (\hat{\mu}_{ij} - k \hat{\sigma}_{ij}, \hat{\mu}_{ij} + k \hat{\sigma}_{ij}) \cap (-c, +c),
	\label{eqn:fij} \\
	\Dcal' &\coloneqq \left\{ j \in [D] : \exists i \in [B] \textrm{ s.t. }  \Fcal_i(j) = \emptyset \right\}
	\label{eqn:kept-feature-set}
\end{align}
%
Intuitively, this means discarding any feature $j$ for which 
$(\hat{\mu}_{ij} - k\hat{\sigma}_{ij}, \hat{\mu}_{ij} + k \hat{\sigma}_{ij})$ overlaps with
$(-c, c)$ for all block indices $i$. If we were to use the Laplace approximation for the posterior $p(W_{ij} | A, X) \approx \Gaussian(W_{ij}; \hat{\mu}_{ij}, \hat{\sigma}_{ij}^2)$, then this is analogous to a hypothesis test on the value of $W_{ij}$ as in~(\ref{eqn:hyp-test-discard});
then $\Dcal'$ comprises all features $j$ for which $H_1$ is accepted at least once for some $i \in [B]$.
%
\begin{equation}
	H_0: |W_{ij}| \leq c \qquad
	H_1: |W_{ij}| > c
	\label{eqn:hyp-test-discard}
\end{equation}
%
The multiplier $k$ in~(\ref{eqn:fij}) determines the degree of significance of the result. However, as the Laplace approximation is not exact, we only treat this dimensionality reduction method as a useful heuristic and not an exact method. Appendix~\ref{appdx:hyp-test} explores this subject at greater depth.

### 2021-06-01 cued/risk.tex
Proof-read night before

As experimental work was entirely computer-based the risks encountered were not what one might typically associate with a more hands-on project: heavy machinery, dangerous chemicals etc. Most of the risks identified would be as a result of an improper working environment. These could be properly managed by ensuring the working environment felt comfortable. An external monitor went some way to alleviating the eye-strain of coding on a small screen for extended periods. Furthermore, an external mouse is far kinder on the wrist than a track-pad on laptops. There were no unexpected hazards encountered that were not identified at the start of the project.

>>>>

As experimental work was entirely computer-based, the risks encountered were not what one might typically associate with a more hands-on project: heavy machinery, dangerous chemicals etc. Most of the risks identified would be as a result of an improper working environment. These could be properly managed by ensuring the working environment felt comfortable. An external monitor went some way to alleviating the eye-strain of coding on a small screen for extended periods. Furthermore, an external mouse is far kinder on the wrist than a track-pad on laptops. There were no unexpected hazards encountered that were not identified at the start of the project.
